---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}
library(ggplot2)
library(dplyr)
getwd()
#download zipped files
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, destfile = "projectdata.zip")
#unzip
unzip(zipfile = "projectdata.zip", exdir = "projectdata")
#check for files in directory
list.files("projectdata")
#read in data files
data <- read.csv("./projectdata/activity.csv")
head(data)
```


## What is mean total number of steps taken per day?

```{r}
#calculate total number of steps per day
steps_per_day <- data %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm = TRUE))

```
## Histogram showing the steps per day

```{r}
#Histogram of steps per day
ggplot(steps_per_day, aes(x = as.Date(date), y = total_steps)) +
    geom_col(fill = "steelblue", color = "black", alpha = 0.7) +
    labs(title = "Total Steps per Day",
         x = "Date",
         y = "Total Steps") +
    scale_x_date(date_breaks = "2 day", date_labels = "%b %d") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))   
```

## The average and median number of steps taken per day are:

```{r}
#mean and median steps per day
mean <- round(mean(steps_per_day$total_steps), 1)
print(paste("The average number of steps is:", mean))
median <- (median(steps_per_day$total_steps))
print(paste("The median number of steps taken is:", median))
```

## What is the average daily activity pattern?

```{r}
#calculate average number of steps per 5 min interval
steps_per_interval <- data %>%
    group_by(interval) %>%
    summarise(interval_steps = mean(steps, na.rm = TRUE))
```

##Time series plot of daily activity pattern:
```{r}
#create time interval plot
ggplot(steps_per_interval, aes(x = interval, y = interval_steps)) +
    geom_line(color = "blue") +
    labs(title = "Average Number of Steps per Interval",
         x = "5-Minute Interval",
         y = "Average Number of Steps") +
    scale_x_continuous(breaks = seq(0, 2355, by = 60)) +  # Adjust the sequence and interval as needed
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## 5-min interval with most steps:

```{r}

max_interval <- steps_per_interval %>%
    filter(interval_steps == max(interval_steps)) %>%
    select(interval, interval_steps)
print(paste("The 5-min interval with the most steps is:", max_interval))  

```


## Imputing missing values

```{r}
#Number of missing values
NAs <- sum(is.na(data$steps))
print(paste("Number of columns with missing values is", NAs))
```
## Strategy to fill missing values
# To fill the missing values we will use the average at each interval and paste it onto the missing values.

```{r}
#Get mean for each interval
steps_per_interval <- data %>%
    group_by(interval) %>%
    summarise(interval_steps = mean(steps, na.rm = TRUE))
```

## New dataset with filled missing values

```{r}
data_filled <- data %>%
    left_join(steps_per_interval, by = "interval") %>%
    mutate(steps = ifelse(is.na(steps), interval_steps, steps)) %>%
    select(-interval_steps)
sum(is.na(data_filled$steps))

data_filled

```

## Histogram and mean and median of dataset with filled values

```{r}

#calculate total number of steps per day
steps_per_day_filled <- data_filled %>%
    group_by(date) %>%
    summarise(total_steps = sum(steps, na.rm = TRUE))
steps_per_day_filled

#Histogram of steps per day
ggplot(steps_per_day_filled, aes(x = as.Date(date), y = total_steps)) +
    geom_col(fill = "red", color = "black", alpha = 0.7) +
    labs(title = "Total Steps per Day",
         x = "Date",
         y = "Total Steps") +
    scale_x_date(date_breaks = "2 day", date_labels = "%b %d") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))    

#mean and median steps per day
mean <- round(mean(steps_per_day$total_steps), 1)
print(paste("The average number of steps is:", mean))
median <- (median(steps_per_day$total_steps))
print(paste("The median number of steps taken is:", median))


```

## Histograms with missing values (top, blue) and with imputed (bottom, red) values

```{r}
ggplot(steps_per_day, aes(x = as.Date(date), y = total_steps)) +
    geom_col(fill = "steelblue", color = "black", alpha = 0.7) +
    labs(title = "Total Steps per Day",
         x = "Date",
         y = "Total Steps") +
    scale_x_date(date_breaks = "2 day", date_labels = "%b %d") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))   

ggplot(steps_per_day_filled, aes(x = as.Date(date), y = total_steps)) +
    geom_col(fill = "red", color = "black", alpha = 0.7) +
    labs(title = "Total Steps per Day",
         x = "Date",
         y = "Total Steps") +
    scale_x_date(date_breaks = "2 day", date_labels = "%b %d") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))   


```


## Are there differences in activity patterns between weekdays and weekends?

```{r}

#Use weedkdays function and add column indicating each weekday
data$date <- as.Date(data$date)
data$day_of_the_week <- weekdays(data$date)
head(data)

#create the "weekend_or_weekday" variable
data$weekday_or_weekend <- ifelse(data$day_of_the_week %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
head(data)
table(data$weekday_or_weekend)

#calculate average number of steps for each 5min interval for weekend and weekday
Average <- data %>%
  group_by(interval, weekday_or_weekend) %>%
  summarize(Average = mean(steps, na.rm = TRUE))

Average

ggplot(Average, aes(x = interval, y = Average)) +
  geom_line() +
  facet_wrap(~weekday_or_weekend, ncol = 1, scales = "free_y") +
  labs(title = "Average Daily Steps by Interval",
       x = "5-Minute Interval",
       y = "Average Number of Steps") +
  theme_minimal()

```
